Galaxy image statistical generator: how to be confident?

### 2024-09-14 paper/main.tex
more experiments

So in the same spirit as \citep{kadkhodaie2024generalization}, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images. The same training procedure used by \cite{kadkhodaie2024generalization} has been applied. Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations. {\color{red} We refer the reader to the  github repository companion for the details of the different codes and notebooks used and models trained as well as the training dataset of $250,000$ galaxies.}
%
\subsection{Results}
%
Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and from a diffusion-based and flow-based models generation using $N=10^5$ trained images. 
%
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-original_samples_ex.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_ex.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
%    \vspace{5mm}
    \begin{subfigure}[t]{0.30\columnwidth}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-glow_samples_ex.pdf}
        %\caption{xxx}
    \end{subfigure}
    \caption{Illustration of the generation of some samples from the diffusion-based (\texttt{U-Net} denoiser) and the \texttt{Glow} flow-based models trained with $N=10^5$ images. On the left some original images are also displayed c-for visual comparison.}
    \label{fig-Original-Glow-UNet-samples}
\end{figure}
At first look the generated images may be not distinguishable from the original images. Notice that we do not expect high resolution images as well as state-of-the-art competitive results, see for instance \citep{ravanbakhsh2016,Fussell2019,Lanusse2021,smith2021,HACKSTEIN2023100685} for better image quality. Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing?

Figure \ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric considering either $1,000$ samples generated by the two diffusion-based models trained with independent dataset of same size $N$ and importantly \textit{seeded by the same white noise images} (top), or considering a single model and finding the closest training image (bottom). Despite the fact that the original training images are less diverse as e.g. the \texttt{CelebA} face or the \texttt{LSUN} bedroom images, we experience the following facts: 
\begin{itemize}
\item first: at low sized dataset ($N=100$), the samples generated by the two models look very similar to a training image (cosine similarity close to 1) as shown by bottom left histogram while the sample generated from the same white noise image looks different as the cosine similarity is different from 1 (upper left histogram);
\item second: greater the dataset size is then more and more generated images looks different (from left to right histograms of the lower row) and simultaneously the samples of the two models are more and more similar (from left to right histograms of the upper row). Already for $N=100,000$ we are close to get a unique bin filled at cosine similarity of 1. 
\end{itemize}

@@

%%%%%%%%%%
\addcontentsline{toc}{section}{References}
% Put your bibiliography file here
%\section{Bibliography}
\bibliographystyle{aa}%{apalike}%{elsarticle-harv}%%{mnras}
\bibliography{refs.bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Details on the dataset}

>>>>

So in the same spirit as \citep{kadkhodaie2024generalization}, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images. The same training procedure used by \cite{kadkhodaie2024generalization} has been applied. Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations. {\color{red} We refer the reader to the  github repository companion for the details of the different codes and notebooks used and models trained as well as the training dataset of $250,000$ galaxies.}
%
\subsection{Results}
%
\subsubsection{Sample images}
Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and from a diffusion-based and flow-based models generation using $N=10^5$ trained images. 
%
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-original_samples_gray.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_gray.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
        \centering
        \raisebox{0mm}{\includegraphics[width=0.8\textwidth]{fig-glow_samples_gray.pdf}}
        %\caption{xxx}
    \end{subfigure}
    \caption{Illustration of the generation of some samples from the diffusion-based (\texttt{U-Net} denoiser) and the \texttt{Glow} flow-based models trained with $N=10^5$ images. On the left some original images are also displayed for visual comparison.}
    \label{fig-Original-Glow-UNet-samples}
\end{figure}
At first look the generated images may be not distinguishable from the original images. Notice that we do not expect high resolution images as well as state-of-the-art competitive results, see for instance \citep{ravanbakhsh2016,Fussell2019,Lanusse2021,smith2021,HACKSTEIN2023100685} for better image quality. Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing?
%
\subsubsection{U-Net denoising performances}
Figures \ref{fig-UNet-denoising-train},\ref{fig-UNet-denoising-test}...
\begin{figure}
    \centering
		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_train.pdf}\\
		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100_denoising_train.pdf}
	\caption{Performances of a some \texttt{U-Net} models using an image of the \textit{training sample}. Top row: (left to right) the original image with progressively a larger noise added. \texttt{PSNR} is defined as $10 \log10$ ratio of squared dynamic range
to mean square error; the distance is the euclidean distance divided between the square root of number of pixels. Middle row: (left to right) (image, PSNR, distance) obtained after denoising the corresponding image of top row with \texttt{U-Net} trained with  $N=100$ images. Bottom row: same as the middle row but   with \texttt{U-Net} trained with  $N=100,000$ images}
	\label{fig-UNet-denoising-train}
\end{figure}

\begin{figure}
    \centering
		\includegraphics[width=0.8\textwidth]{fig-UNet_100000_denoising_test.pdf}\\
		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_test.pdf}
	\caption{Similar to  Figure \ref{fig-UNet-denoising-train} but  using an image never used to train any models.}
	\label{fig-UNet-denoising-test}
\end{figure}
%
\subsubsection{Cosine similarity metric}
Figure \ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric considering either $1,000$ samples generated by the two diffusion-based models trained with independent dataset of same size $N$ and importantly \textit{seeded by the same white noise images} (top), or considering a single model and finding the closest training image (bottom). Despite the fact that the original training images are less diverse as e.g. the \texttt{CelebA} face or the \texttt{LSUN} bedroom images, we experience the following facts: 
\begin{itemize}
\item first: at low sized dataset ($N=100$), the samples generated by the two models look very similar to a training image (cosine similarity close to 1) as shown by bottom left histogram while the sample generated from the same white noise image looks different as the cosine similarity is different from 1 (upper left histogram);
\item second: greater the dataset size is then more and more generated images looks different (from left to right histograms of the lower row) and simultaneously the samples of the two models are more and more similar (from left to right histograms of the upper row). Already for $N=100,000$ we are close to get a unique bin filled at cosine similarity of 1. 
\end{itemize}

@@

%%%%%%%%%%
\addcontentsline{toc}{section}{References}
% Put your bibiliography file here
%\section{Bibliography}
\bibliographystyle{aa}%{apalike}%{elsarticle-harv}%%{mnras}
\bibliography{refs.bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\appendix
%\section{}
